When does vapor pressure deficit drive or reduce evapotranspiration?

### 2018-01-05 doc/paper/vpd_et_paper.tex
finished with my moring writing, need to revert src back to previous commit though for figures to be good

Answering the question ``When does vapor pressure deficit drive or reduce evapotranspiration?'' is motivated by two potenential but opposing perspectives on the matter. The first persepctive is that increases to vapor pressure deficit (VPD) increases atmospheric demand for water from the land surface, and this drives an increase in evapotranspiration (ET). This perspective is particularly relevant because potential evapotranspiration (PET), which is used in many drought indicies and hydrometeorological studies, only quantifies this atmospheric demand perspective. However, there is another possible perspective for vegetated surfaces. Plants' stomata have evolved to optimally regulate the exchange of water and carbon, and tend to close in response to increased atmospheric dryness \citep{Ball_1987, Leuning_1990, MEDLYN_2011}.  Therefore, an increase in VPD may actually correspond to a decrease in ET because stomatal closure can cancel the effects of shifts to atmospheric de. So really a question of ``When does VPD drive or reduce ET?'' becomes a question of ``When does plant response or atmospheric demand dominate?.''

Quantifying the plant response to a perturbation to atmospheric VPD increases our understanding of land surface response to shifts in atmospheric conditions. If plant response reduces ET in response to atmospheric drying then soil moisture will be conserved. An increase in ET in reponse to atmospheric drying will reduce soil moisture, but contribute increased moisenting to the atmosperhere. Understanding the one way response of the land surface to the atmosphere is a critical first step to disentangling land-atmosphere feedbacks. The sign and magnitude of land-surface response drives the co-evolution of the atmosphere and land-surface at many timescales, from diurnal to interdecadal.

Before diving into math and data, we have some prior intuition for how the system might behave. First, whether VPD drive or reduces ET should be a function of plant type. Plants that evolved to conserve water (e.g. arid shrubs) will be more likely to reduce ET with increasing VPD, and plants that have evolved to care little about water (e.g. crops) will be more likely to increase ET with increasing VPD. Additionally, atmospheric conditions must matter as well. At the ecosystem scale there are limits to the strategies plants use to hold water back from the atmosphere. As atmospheric demand for water (VPD) increases, ecosystems will begin to reach their water conservation limits. At this stage any further increase in VPD will most likely drive an increase in ET, because there is little more the plants can do keep water back.

@@

\section{Data}
\label{data}
We use data from FLUXNET2015, including all sites with more than four years of data. Each site's plant functional type (PFT) was classified using the International Geosphere-Biosphere Programme vegetation classification scheme (\cite{Loveland_1999}).  Because $g_1$ coefficients \citep{Lin_2015} and uWUE were only both available for five plant functional types (PFTs - see Table \ref{pft}),  only 56 of the 77 sites were used. Figure \ref{map_fig} \explain{map needs to be improved - it's a placeholder for now} presents each site and its plant functional type.

The purpose of this study is examine ecosystem response to atmospheric drying during the growing season. To accomplish this, we filter and quality control the data using a similiar procedure as \cite{Zhou_2015}:
\begin{itemize}
\item Only measured or highest (``good'') quality gapfilled data, according to qualtiy control flags, are used.
\item To isolate the growing season, we only use days in which the average GPP exceeds 10\% of the observed 95th percentile of GPP for a given site.
\item We remove days with rain and the day following to avoid issues with rain interveption and sensor satruation and high relative humidity (Medlyn et. al. 2017).
\end{itemize}
Additionally, we restrict data to the daytime which is identified when downwelling shortwave radation is greater than 50 W m$^{-2}$ and sensible heat flux is greater than 5 W m$^{-2}$. To further reduce the chance of sensor saturation at high relative humidity, we remove all time steps for which VPD is less than .01 kPa. Timesteps with negative observed GPP or ET are also removed. After these quality control proceedures, 332556 upscaled hourly observations remain.

\section{Methods}
The Penman-Monteith equation (hereafter PM) estimates ET as a function of atmospheric and land-surface variables:
\begin{linenomath*}
  \begin{equation}
      ET = \frac{\Delta R + g_a \rho_a c_p VPD}{\Delta + \gamma(1 + \frac{g_a}{g_s})},
  \end{equation}
\end{linenomath*}

@@

\begin{figure}[h]
\centering
\includegraphics[width=20pc]{./fig01.pdf}
\caption{Plant functional type and location of sites used in analysis. ***This is just a placeholder for now and needs to be improved i.e. with lat lon, better placement of continents, etc.)***}
\label{map_fig}
 \end{figure}

@@

By construction, the variability in the $\sigma$ term (Equation \ref{lai}) contains all model and observational uncertainties. For an observation that perfectly matches our model and constant uWUE assumption $\sigma$ will be one. Therefore, for our assumptions and framework to be reasonable $\sigma$ should be $O(1)$. Figure \ref{lai_fig} presents the histogram of calculated $\sigma$s with outliers (lowest and highest 5\% percent). All remaining $\sigma$ values are $O(1)$ which provides confidence in model framework.

@@

% below show histogram of site specific spearmannr? think would be useful 
\begin{figure}[h]
\centering
\includegraphics[width=20pc]{./fig03.png}
\caption{The joint distribution of $VPD$ and $\sigma$. $\sigma$ has only a weak dependence on $VPD$. ***This plot could proabably benefit from a box plot of site specific correlations, because some sites do have stronger depednence than others. Note also Figs 3 and 2 can probably be combined because this figure shoes $\sigma$'s histogram.***}
\label{lai_vpd_fig}
\end{figure}nnn

Before calculating the sensitivity of ET to VPD, it is useful to consider the functional form of Equation \ref{d_et}. There are two main terms: a ``scaling'' term which modifies the mangnitude but not the sign:

\[\frac{g_a \; P}{T(\Delta + \gamma)},\]

@@

\[\frac{c_p}{R_{air}} - \frac{\gamma c_s }{1.6 \; R\; \text{ uWUE }} \left( \frac{2 g_1 + \sqrt{VPD}}{2 (g_1 + \sqrt{VPD})^2}\right).\]
 All variables are positive, so the relative magnitude between the first term and the second term will deterimine whether ET increases or decreases with increasing VPD. If the second term is larger magnitude then plant response dominates, while if the first term is larger then atmospheric demand dominats. Given that the sign term is crucial to answering our question ``When does VPD drive or reduce ET?'', we will explore its functional form first.

@@

Term 2 minus Term 3's role in determining the sign of the sensitivty of ET to VPD makes it crucuial for answering our question ``When does VPD drive or reduce ET?'' Exploring these terms more, $c_s$ and $\gamma$ variability is relatively less than $\sigma$ and $VPD$ variability, so variability within PFT will be solely determined by $\sigma$ and $VPD$. If we fix uncertainty $\sigma$ at 1, then Term 2 minus Term 3 is just a function of $VPD$. We can further dertmine a $VPD_{crit}$ where $\frac{\partial \; ET}{\partial \; VPD} = 0$:

@@

Figure \ref{term3} provides greater description for how the sign term varies with $VPD$, as a function of PFT. Equation \ref{d_et_approx} aids interpretation of Figure \ref{term3}. Larger $uWUE$, and $g1$ shift the leading-order constant term ($\frac{1}{g_1}$) towards smaller values, and the sign term towards overall positive values. $uWUE$ and $g1$ are both water-use effiency type constants. Higher values corresponde to plants that are more willing to spend water on primary production and less evolved to conserve water.\explain{really need to think about this sentence more, tehre is more nuance and I actually thing g1 and uwue really partially cancel.} Figure \ref{term3} confirms our physical intuition: CROs are the least water conservative so have the smallest constant portion of Term 3, while CSH are the most water conservative and have the largest constant portion of Term 3. For the VPD-dependent terms in Equation \ref{d_et_approx}, differences in $g1$ between PFTs exert a greater influence than difference $uWUE$, as the power of $g1$ increases. Increasing $uWUE$ and $g1$ decreases the VPD-dependence, but $g1$ has the bigger effect due to its increasing powers. ENF ($g1 = 74.31$) has by far the largest VPD dependence of response, while CRO ($g1 = 183.1$) has the smallest VPD dependence.\explain{could probably simplify and just say that g1 controls VPD dependence.}

\begin{figure}[h]
\centering
\includegraphics[width=20pc]{./fig05.pdf}
\caption{Sources of variability for sign term. Top: sign term as a function of VPD, with $\sigma$ held fixed at PFT averages. The observed range of VPD for each PFT is also shown below the x-axis. Line extent corresponds to 5th and 95th percentiles, while stars denote the location of the 25th, 50th, and 75th percentiles.

Bottom: The location of the minima of ET, as a function of VPD and $\sigma$. Lines and stars denote the distribution of VPD and $\sigma$ next to each axis, following the same percentiles as above.}
\label{term3}
\end{figure}

Figure \ref{term3}b shows the location of the minima of ET, as a function of $\sigma$ and $VPD$. For any $\sigma$ or VPD less (more) than these curves, sign term will be negative (positive). It is clear that the portion of VPD observations below/above these curves will be a strong function of $\sigma$. However, we can see some general trends. For CSH, $\frac{\partial \; ET}{\partial \; VPD}$ should be negative for the vast majority of observed $\sigma$ and VPD. The fraction of positive $\frac{\partial \; ET}{\partial \; VPD}$ appears to be more even for ENF, GRA, and DBF, and we might expect a greater frequency of positive $\frac{\partial \; ET}{\partial \; VPD}$ for CRO.

\subsection{Functional Form Scaling Term}

While the above discussion of the sign of $\frac{\partial \; ET}{\partial \; VPD}$ is important to answer our research question, the magnitude of $\frac{\partial \; ET}{\partial \; VPD}$ will also impact statistics of $\frac{\partial \; ET}{\partial \; VPD}$ and the importance of $VPD$ variablility for overall $ET$ variability. So we will more closely examine the scaling Term 1: $\frac{P}{T} \propto \rho$, so this should vary little relative to aerodynamic conductance and $\Delta$. $\gamma$ should also be relatively constant, so Term 1 should be primarily a function of aerodynamic conductance and temperature (through the function $\Delta$). This makes sense, as aerodynamic conductance represents how efficently response at the surface is communicated to the atmosphere. As it increases, any plant response will be communicated more strongly to the atmosphere (and vice-versa).

@@

Figure \ref{scale_vary}A shows scaling term normalized by mean aerodynamic conductance (calculated for each plant functional type), and confirms that much of the relative variability of the scaling term is contained in the aerodynamic conductance variability. Generally, $T$ has less of a role. Additionally, the impact of $T$ on the relative variability increases with increasing aerodynamic conductance.

While the relative variability of scaling term is similar across PFT, the absolute value of scaling term varies strongly across PFT. Figure \ref{scale_vary}B shows scaling term evaluated with the mean aerodynamic conductance for each PFT, and at the range of observed temperatures for each PFT. As expected, for the tree PFTs (DBF, ENF) scaling term is much larger and the temperature dependence is much stronger. Systematic differences in observed temperatures also cause differences in the average magnitude of scaling term. For example, ENF experiences on average colder temperatures and is thus more likely to have a larger scaling term. Additionally, because the variability of aerodynamic conductance increases proportionally to the mean, the spread of scaling term due to aerodynamic conductance variability will be larger for the tree PFTs, although this is not shown for simplicity. To summarize, the variability of scaling term within each PFT will look like Figure \ref{scale_vary}A for each PFT, but the scale of the y-axis will increase or decrease according to mean aerodynamic conductance oberved in Figure \ref{scale_vary}B.

@@

The units of $\frac{\partial \; ET}{\partial \; VPD}$ make it difficult to interpret if $VPD$ is even a meaningful contributor to ET's variability. To better understand $VPD$'s contribution, we mulitply $\frac{\partial \; ET}{\partial \; VPD}$ with $VPD$'s standard deviation to define a (linearized) relative change in ET for variations in $VPD$ . $VPD$'s contribution to ET's variability ranges between 30 - 40 W m$^{-2}$ for all PFTs except for CSH, which is about 100 W m$^{-2}$. Another meaningful comparison is to $\frac{\partial \; ET}{\partial \; R} \cdot std(R)$, as net radiation is generally the driver of ET (cite joe berry here). For all PFTs except for CSH $VPD$ contributes between 30 - 40 \% of $R$'s contribution to variability. For CSH the portion is much larger, about 88 \%. $VPD$'s variability is certiantly a primary contributor to $ET$'s variability.

So far, idealized plots and statistics have illuminated the form of $\frac{\partial \; ET}{\partial \; VPD}$ and how it varies with PFT. Large mean uWUE shifts CRO and DBF towards positive $\frac{\partial \; ET}{\partial \; VPD}$. However, the strongly nonlinear function of $\frac{\partial \; ET}{\partial \; VPD}$ at $\frac{\partial \; ET}{\partial \; VPD} < 0$ pushes $\overline{\frac{\partial \; ET}{\partial \; VPD}}$ negative for DBF (it does not do this for CRO because of CRO's high $g1$). ENF's low $g1$ value increases the dependence of $\frac{\partial \; ET}{\partial \; VPD}$ on $VPD$, and makes the function more strongly nonlinear. This has the side effect of pushing $\overline{\frac{\partial \; ET}{\partial \; VPD}}$ negative further than other PFTs for a given fraction $\frac{\partial \; ET}{\partial \; VPD} < 0$ and magnitude $\frac{\partial \; ET}{\partial \; VPD}(\overline{T,\ldots,VPD})$. GRA shows the opposite behavior; a relatively high $g1$  makes the function more linear, decreasing the magnitude of $-\overline{\frac{\partial \; ET}{\partial \; VPD}}$ for a given  [large] fraction $\frac{\partial \; ET}{\partial \; VPD} < 0$ and negative $\frac{\partial \; ET}{\partial \; VPD}(\overline{T,\ldots,VPD})$, However, aerodynamic conductance and scaling term also probably have a role in this. Finally, the low $uWUE$ of CSH pushes it toward by far the lowest values $\frac{\partial \; ET}{\partial \; VPD}$ (Figure \ref{term3}). Variability in $VPD$ accounts for the largest about of $ET$ variability for CSH. For the other PFTs, $VPD$ contributes less to $ET$ variability, but still represents about 15-20\explain{depending on how we take the derivative, this could be 30-40\%.} \% of $R$'s contributions to ET variability.

\begin{table}
\caption{Statistics of $\frac{\partial \; ET}{\partial \; VPD}$ as a function of PFT.}
\centering
\begin{tabular}{l c c c c c}
  \hline
PFT & $\overline{\frac{\partial \; ET}{\partial \; VPD}}$ & $\frac{\partial \; ET}{\partial \; VPD}\left(\overline{env}\right)$ & $\frac{\partial \; ET}{\partial \; VPD}\left(\overline{env}\right)*\text{std}(VPD)$ & $\frac{\frac{\partial \; ET}{\partial \; VPD}\left(\overline{env}\right)*\text{std}(VPD)}{ \frac{\partial \; ET}{\partial \; R}\left(\overline{env}\right)*\text{std}(R)}$ & fraction $\frac{\partial \; ET}{\partial \; VPD} < 0.$ \\
  \hline
CRO & 0.000853 & 0.026241 & 37.05 & 0.41 & 0.473311\\
CSH & -0.108234 & -0.091526 & 101.72 & 0.88 & 0.931660\\
DBF & -0.012727 & 0.013794 & 39.47 & 0.33 & 0.461674\\
ENF & -0.034087 & 0.000706 & 33.22 & 0.30 & 0.534425\\
GRA & -0.019637 & -0.000921 & 33.60 & 0.35 & 0.631735\\
\hline
\multicolumn{2}{l}{$^{a}$Footnote text here.}

@@

ENF is a special case. The VPD dependence of the theory is not well captured by observations. However, the theory closely captures the intercept where $\frac{\partial \; ET}{\partial \; VPD} = 0$. Really resolving this feature is the key to answering our research question, so perhaps the theory is still useful for ENF for the purpose of addressing when VPD drives or reduces ET. It is also noteworthy that we used published constants for $g1$, which determines the VPD dependence, from \cite{Lin_2015}. We could have fit our own constants to tweak the theory to match the observations, but this was not our goal.

@@

Figure \ref{real} presents calculated $\frac{\partial \; ET}{\partial \; VPD}$ where, unless otherwise noted, all variables in Equation \ref{d_et} are allowed to vary, including uncertainty. Each column is a different quantity related to $\frac{\partial \; ET}{\partial \; VPD}$, and each row is a different PFT. Vertical black lines denote where our theory says $VPD_{crit}$ should be. In addition to seeing how the theory $VPD_{crit}$ matches observations, these plots are also useful for observing how the scaling term affects the magnitude of the derivative, if at all.

The scatter plots generally confirm expectations from Figure \ref{agu_fig}. By comparing different columns of Figure \ref{real} we can see the impact of $\sigma$ and $g_a$ on the variability of $\frac{\partial \; ET}{\partial \; VPD}$. $g_a$'s scaling (included in columns 1 and 3) alters the magnitude considerably, particularly begtween observations. $\sigma$ variability (included in columns 1 and 2) adds a lot of additional noise to the signal of $\frac{\partial \; ET}{\partial \; VPD}$, as observed in Figure \ref{agu_fig}. However, the general story with the noise appears to match the cleaner signal when $\sigma$ is help constant and $VPD_{crit}$ is clearly visible. The exceptions are, as discussed above, for GRA and CRO, where the full complexity plots (Columns 1 and 2) to not match well with $\sigma$ held fixed (Columns 3 and 4).

While $g_a$ has a large impact on magnitudes in the scaling, temperature, through the $\Delta$ term, does not appear to have as much of an impact in the full observations. Therefore it appears that the absolute magnitude of the derivative can be mostly attributed to the sign term, and its direct scaling by aerodynamic conductance. Intuitively this makes sense; aerodynamic conductance will control how dominant balances at the land surface are communicated to the boundary layer. While $\Delta$ controlls the efficiency of latent fluxes in carying heat away from the surface, it appears this is a second order term, relative to $g_a$, for scaling $\frac{\partial \; ET}{\partial \; VPD}$.

In general Figures \ref{real} and \ref{agu_fig} match our expectations, even with the large sensitivity to $\sigma$ seen in Figure \ref{term3}. Our $\sigma$-based method of uncertainty propagation blurs the idealized expectations the most for GRA and CRO. We therefor have the most confidence in our conclusion based on Equation \ref{d_et} for PFTS CSH, DBF, and ENF, as the sign of the derivative with uncertainty included matches (to leading order) the story when $\sigma$ is held fixed.

>>>>

Answering the question ``When does vapor pressure deficit drive or reduce evapotranspiration?'' is motivated by two potenential but opposing perspectives on the matter. The first persepctive is that increases to vapor pressure deficit (VPD) increases atmospheric demand for water from the land surface, and this drives an increase in evapotranspiration (ET). This perspective is particularly relevant because potential evapotranspiration (PET), which is used in many drought indicies and hydrometeorological studies, only quantifies atmospheric demand effects. However, there is another possible perspective for vegetated surfaces. Plants' stomata have evolved to optimally regulate the exchange of water and carbon, and tend to close in response to increased atmospheric dryness \citep{Ball_1987, Leuning_1990, MEDLYN_2011}.  Therefore, an increase in VPD may actually correspond to a decrease in ET because stomatal closure can cancel the effects of shifts to atmospheric de. So really a question of ``When does VPD drive or reduce ET?'' becomes a question of ``When does plant response or atmospheric demand dominate?.''

Quantifying the plant response to a perturbation to atmospheric VPD increases our understanding of land surface response to shifts in atmospheric conditions. If plant response reduces ET in response to atmospheric drying then soil moisture will be conserved for later ET. An increase in ET in reponse to atmospheric drying will reduce soil moisture, but contribute increased moisenting to the atmosperhere. Understanding the one way response of the land surface to the atmosphere is a critical first step to disentangling land-atmosphere feedbacks. The sign and magnitude of land-surface response drives the co-evolution of the atmosphere and land-surface at many timescales, from diurnal to interdecadal.

Before diving into math and data, we have some prior intuition for how the system might behave. First, whether VPD drives or reduces ET should be a function of plant type. Plants that evolved to conserve water (e.g. arid shrubs) will be more likely to reduce ET with increasing VPD, and plants that have evolved to care little about water (e.g. crops) will be more likely to increase ET with increasing VPD. Additionally, atmospheric conditions must matter as well. At the ecosystem scale there are limits to the strategies plants use to hold water back from the atmosphere. As atmospheric demand for water (VPD) increases, ecosystems will begin to reach their water conservation limits. At this stage any further increase in VPD will most likely drive an increase in ET, because there is little more the plants can do keep water back.

@@

\section{Data}
\label{data}
We use data from FLUXNET2015, including all sites with more than four years of data. Each site's plant functional type (PFT) was classified using the International Geosphere-Biosphere Programme vegetation classification scheme (\cite{Loveland_1999}).  Because physical constats used in Section \ref{methods} are only published for five plant functional types (PFTs - see Table \ref{pft}),  only 56 of the 77 sites were used. Figure \ref{map_fig} \explain{map needs to be improved - it's a placeholder for now} presents each site and its plant functional type.

The purpose of this study is to examine ecosystem response to atmospheric drying during the growing season. To accomplish this, we filter and quality control the data using a similiar procedure as \cite{Zhou_2015}:
\begin{itemize}
\item Only measured or highest (``good'') quality gapfilled data, according to qualtiy control flags, are used.
\item To isolate the growing season, we only use days in which the average GPP exceeds 10\% of the observed 95th percentile of GPP for a given site.
\item We remove days with rain and the day following to avoid issues with rain interveption and sensor satruation and high relative humidity (Medlyn et. al. 2017).
\end{itemize}
Additionally, we restrict data to the daytime which is identified when downwelling shortwave radation is greater than 50 W m$^{-2}$ and sensible heat flux is greater than 5 W m$^{-2}$. To further reduce the chance of sensor saturation at high relative humidity, we remove all time steps for which VPD is less than .01 kPa. Timesteps with negative observed GPP or ET are also removed. After these quality control proceedures, 332556 upscaled hourly observations remain.

\section{Methods}
\label{methods}
The Penman-Monteith equation (hereafter PM) estimates ET as a function of atmospheric and land-surface variables:
\begin{linenomath*}
  \begin{equation}
      ET = \frac{\Delta R + g_a \rho_a c_p VPD}{\Delta + \gamma(1 + \frac{g_a}{g_s})},
  \end{equation}
\end{linenomath*}

@@

\begin{figure}[h]
\centering
\includegraphics[width=20pc]{./fig01.pdf}
\caption{Plant functional type and location of sites used in analysis. ***\textbf{This is just  aplaceholder for now, I am currently fixing it***}}
\label{map_fig}
 \end{figure}

@@

By construction, the variability in the $\sigma$ term (Equation \ref{lai}) contains all model and observational uncertainties. For an observation that perfectly matches our model and constant uWUE assumption $\sigma$ will be one. Therefore, for our assumptions and framework to be reasonable $\sigma$ should be $O(1)$. Figure \ref{lai_fig} presents the histogram of calculated $\sigma$s with outliers removed (lowest and highest 5\% percent). All remaining $\sigma$ values are $O(1)$ which provides confidence in model framework.

@@

% below show histogram of site specific spearmannr? think would be useful 
\begin{figure}[h]
\centering
\includegraphics[width=20pc]{./fig03.png}
\caption{The joint distribution of $VPD$ and $\sigma$. $\sigma$ has only a weak dependence on $VPD$. \textbf{***note I am redoing/cleaning up this plot, ``lai'' should read sigma***}}
\label{lai_vpd_fig}
\end{figure}

Before calculating the sensitivity of ET to VPD, we will consider the functional form of Equation \ref{d_et}. There are two main terms: a ``scaling'' term which modifies the mangnitude but not the sign:

\begin{equation}
  \frac{g_a \; P}{T(\Delta + \gamma)},
\end{equation}

@@

\begin{equation}
  \label{sign}
  \frac{c_p}{R_{air}} - \frac{\gamma c_s }{1.6 \; R\; \text{ uWUE }} \left( \frac{2 g_1 + \sqrt{VPD}}{2 (g_1 + \sqrt{VPD})^2}\right).
\end{equation}
All variables are positive, so the relative magnitude between the first term and the second term in Equation \ref{sign} will deterimine whether ET increases or decreases with increasing VPD. If the second term is larger magnitude then plant response dominates, while if the first term is larger then atmospheric demand dominats. Given that the sign term is crucial to answering our question ``When does VPD drive or reduce ET?'', we will explore its functional form first.

@@

The sign term is crucuial for answering our question ``When does VPD drive or reduce ET?,'' so let's explore the terms within it more. The variability of $c_s$ and $\gamma$ is relatively less than $\sigma$ and $VPD$ variability, so we sign term variability within PFT will approximately determined by just $\sigma$ and $VPD$. If we fix uncertainty $\sigma$ at 1, then the sign term is approximately just a function of $VPD$. We can further dertmine a $VPD_{crit}$ where $\frac{\partial \; ET}{\partial \; VPD} = 0$:

@@

Figure \ref{term3} provides greater description for how the sign term varies with $VPD$, as a function of PFT. Equation \ref{d_et_approx} aids interpretation of Figure \ref{term3}. Larger $uWUE$ and $g1$ shift the leading-order constant term ($\frac{1}{g_1}$) towards smaller values, and the sign term towards overall positive values. % Higher values corresponde to plants that are more willing to spend water on primary production and less evolved to conserve water.\explain{really need to think about this sentence more, tehre is more nuance and I actually thing g1 and uwue really partially cancel.}
Figure \ref{term3} confirms our physical intuition: CROs are the least water conservative so have the smallest constant portion of Term 3, while CSH are the most water conservative and have the largest constant portion of Term 3. For the VPD-dependent terms in Equation \ref{d_et_approx}, differences in $g1$ between PFTs exert a greater influence than difference $uWUE$, as the power of $g1$ increases. Increasing $uWUE$ and $g1$ decreases the VPD-dependence, but $g1$ has the bigger effect due to its increasing powers. ENF ($g1 = 74.31$) has by far the largest VPD dependence of response, while CRO ($g1 = 183.1$) has the smallest VPD dependence.

\begin{figure}[h]
\centering
\includegraphics[width=20pc]{./fig05.pdf}
\caption{Sources of variability for the sign term. Top: sign term as a function of VPD, with $\sigma$ held fixed at 1. The observed range of VPD for each PFT is also shown below the x-axis. Line extent corresponds to 5th and 95th percentiles, while stars denote the location of the 25th, 50th, and 75th percentiles.

Bottom: The location of the minima of ET, as a function of VPD and $\sigma$. Lines and stars denote the distribution of VPD and $\sigma$ next to each axis, following the same percentiles as above. \textbf{***need to replot with correct sigma (normalzied to 1)***}}
\label{term3}
\end{figure}

Figure \ref{term3}b shows the location of the minima of ET, as a function of $\sigma$ and $VPD$. For any $\sigma$ or VPD less (more) than these curves, the sign term will be negative (positive). It is clear that the portion of VPD observations below/above these curves will be a strong function of $\sigma$. However, we can see some general trends. For CSH, $\frac{\partial \; ET}{\partial \; VPD}$ should be negative for the vast majority of observed $\sigma$ and VPD. The fraction of positive $\frac{\partial \; ET}{\partial \; VPD}$ appears to be more even for ENF, GRA, and DBF, and we might expect a greater frequency of positive $\frac{\partial \; ET}{\partial \; VPD}$ for CRO.

\subsection{Functional Form of the Scaling Term}

While the above discussion of the sign of $\frac{\partial \; ET}{\partial \; VPD}$ is important to answer our research question, the magnitude of $\frac{\partial \; ET}{\partial \; VPD}$ will also impact statistics of $\frac{\partial \; ET}{\partial \; VPD}$ and the importance of $VPD$ variablility for overall $ET$ variability. So we will more closely examine the scaling term: $\frac{P}{T} \propto \rho$, so this should vary little relative to aerodynamic conductance and $\Delta$. $\gamma$ should also be relatively constant, so the scaling term should be primarily a function of aerodynamic conductance and temperature (through the function $\Delta$). This makes sense, as aerodynamic conductance represents how efficently response at the surface is communicated to the atmosphere. As it increases, any plant response will be communicated more strongly to the atmosphere (and vice-versa).

@@

Figure \ref{scale_vary}A shows scaling term normalized by mean aerodynamic conductance (calculated for each plant functional type), and confirms that much of the relative variability of the scaling term is caused by aerodynamic conductance variability. Generally, temperature (via $\Delta$) causes less relative variability. However, the impact of $T$ on the relative variability increases with increasing aerodynamic conductance.

While the relative variability of scaling term is similar across PFT, the absolute value of scaling term varies strongly across PFT. Figure \ref{scale_vary}B shows scaling term evaluated with the mean aerodynamic conductance for each PFT, and at the range of observed temperatures for each PFT. As expected, for the tree PFTs (DBF, ENF) the scaling term is much larger and the temperature dependence is much stronger. Systematic differences in observed temperatures also cause differences in the average magnitude of scaling term. For example, ENF experiences on average colder temperatures and is thus more likely to have a larger scaling term. Additionally, because the variability of aerodynamic conductance increases proportionally to the mean, the spread of the scaling term due to aerodynamic conductance variability will be larger for the tree PFTs, although this is not shown for simplicity. To summarize Figure \ref{scale_vary}, the variability of the scaling term within each PFT will look like Figure \ref{scale_vary}A for each PFT, but the scale of the y-axis will increase or decrease according to mean aerodynamic conductance oberved in Figure \ref{scale_vary}B.

@@

The units of $\frac{\partial \; ET}{\partial \; VPD}$ make it difficult to interpret if $VPD$ is even a meaningful contributor to ET's variability. To better understand $VPD$'s contribution, we mulitply $\frac{\partial \; ET}{\partial \; VPD}$ with $VPD$'s standard deviation to define a (linearized) relative change in ET for variations in $VPD$ . $VPD$'s contribution to ET's variability ranges between 30 - 40 W m$^{-2}$ for all PFTs except for CSH, which is about 100 W m$^{-2}$. Another meaningful comparison is to $\frac{\partial \; ET}{\partial \; R} \cdot std(R)$, as net radiation is generally the driver of ET (cite joe berry here)\explain{Find citation?}. For all PFTs except for CSH $VPD$ contributes between 30 - 40 \% of $R$'s contribution to variability. For CSH the portion is much larger, about 88 \%. $VPD$'s variability is certiantly a non-negligable contributor to $ET$'s variability.

So far, idealized plots and statistics have illuminated the form of $\frac{\partial \; ET}{\partial \; VPD}$ and how it varies with PFT. Large mean uWUE shifts CRO and DBF towards positive $\frac{\partial \; ET}{\partial \; VPD}$, but $\overline{\frac{\partial \; ET}{\partial \; VPD}}$ is still negative for DBF due to its lower $g1$. ENF's low $g1$, even lower than DBF, increases the dependence of $\frac{\partial \; ET}{\partial \; VPD}$ on $VPD$, and makes the function very nonlinear. This has the side effect of pushing $\overline{\frac{\partial \; ET}{\partial \; VPD}}$ negative further than other PFTs for a given fraction $\frac{\partial \; ET}{\partial \; VPD} < 0$ and magnitude $\frac{\partial \; ET}{\partial \; VPD}(\overline{T,\ldots,VPD})$. GRA shows the opposite behavior; a relatively high $g1$  makes the function more linear, decreasing the magnitude of $\overline{\frac{\partial \; ET}{\partial \; VPD}}$ for a given fraction $\frac{\partial \; ET}{\partial \; VPD} < 0$ and negative $\frac{\partial \; ET}{\partial \; VPD}(\overline{T,\ldots,VPD})$. Finally, the low $uWUE$ of CSH pushes it toward by far the lowest values $\frac{\partial \; ET}{\partial \; VPD}$ (Figure \ref{term3}). Variability in $VPD$ accounts for the largest about of $ET$ variability for CSH. For the other PFTs, $VPD$ contributes less to $ET$ variability, but still represents about 30-40 \% of $R$'s contributions to ET variability.

\begin{table}
\caption{Statistics of $\frac{\partial \; ET}{\partial \; VPD}$ as a function of PFT.}
\centering
\begin{tabular}{l c c c c c}
  \hline
PFT & $\overline{\frac{\partial \; ET}{\partial \; VPD}}$ & $\frac{\partial \; ET}{\partial \; VPD}\left(\overline{env}\right)$ & $\frac{\partial \; ET}{\partial \; VPD}\left(\overline{env}\right)*\text{std}(VPD)$ & $\frac{\frac{\partial \; ET}{\partial \; VPD}\left(\overline{env}\right)*\text{std}(VPD)}{ \frac{\partial \; ET}{\partial \; R}\left(\overline{env}\right)*\text{std}(R)}$ & fraction $\frac{\partial \; ET}{\partial \; VPD} < 0.$ \\
  \hline
CRO & 0.000853 & 0.026241 & 37.05 & 0.41 & 0.473311\\
CSH & -0.108234 & -0.091526 & 101.72 & 0.88 & 0.931660\\
DBF & -0.012727 & 0.013794 & 39.47 & 0.33 & 0.461674\\
ENF & -0.034087 & 0.000706 & 33.22 & 0.30 & 0.534425\\
GRA & -0.019637 & -0.000921 & 33.60 & 0.35 & 0.631735\\
\hline

@@

ENF is a special case. The VPD dependence of the theory is not well captured by observations. However, the theory closely captures the intercept where $\frac{\partial \; ET}{\partial \; VPD} = 0$. Resolving this feature is the key to answering our question, so perhaps the theory is still useful for ENF for the purpose of addressing when VPD drives or reduces ET. It is also noteworthy that we used published constants for $g1$, which determines the VPD dependence, from \cite{Lin_2015}. We could have fit our own constants to tweak the theory to match the observations, but our goal was to use a priori physical constants whenever possible..

@@

A final comparison for Figure \ref{agu_fig} is to the value of the sign term for PET. For PET, the sign term is a constant $\frac{c_p}{R_{air}}\approx 3.5$. This is far more than any of the observed values in Figure \ref{agu_fig}, even when uncertainty is included. So, even when atmospheric demand dominates plant response, the plant response still has a large role in attenuating atmospheric demand. In order to understand and capture the behavior of land-surface reponse, representation of the plants must be included in some form. This comparison casts doubt on the usefulness of PET for interprenting land-atmosphere evolution, as is done often in hydrometeorlogical studies.

Figure \ref{real} presents calculated $\frac{\partial \; ET}{\partial \; VPD}$ where, unless otherwise noted, all variables in Equation \ref{d_et} are allowed to vary, including uncertainty. Each column is a different quantity related to $\frac{\partial \; ET}{\partial \; VPD}$, and each row is a different PFT. Vertical black lines denote where our theory says $VPD_{crit}$ should be. For any obervations to the left of this line $\frac{]aprial \; ET}{\partial \; VPD}$ should be negative, and any ovservations the righ right of the line should be postiive. In addition to seeing how the theory $VPD_{crit}$ matches observations, these plots are also useful for observing how the scaling term affects the magnitude of the derivative, if at all.

The scatter plots generally confirm expectations from Figure \ref{agu_fig}. By comparing different columns of Figure \ref{real} we can see the impact of $\sigma$ and $g_a$ on the variability of $\frac{\partial \; ET}{\partial \; VPD}$. $g_a$'s scaling (included in columns 1 and 3) alters the magnitude considerably, particularly between PFTs. $\sigma$ variability (included in columns 1 and 2) adds a lot of additional noise to the signal of $\frac{\partial \; ET}{\partial \; VPD}$, as observed in Figure \ref{agu_fig}. However, the general story with the noise appears to match the cleaner signal when $\sigma$ is help constant and $VPD_{crit}$ is clearly visible. The exceptions are, as discussed above, for GRA and CRO, where the full complexity plots (Columns 1 and 2) to not match well with $\sigma$ held fixed (Columns 3 and 4).

While $g_a$ has a large impact on magnitudes in the scaling, temperature, through the $\Delta$ term, does not appear to have as much of an impact in the complete estimation of $\frac{\partial \; ET}{\partial \; VPD}$. Therefore it appears that the absolute magnitude of the derivative can be mostly attributed to the sign term, and its direct scaling by aerodynamic conductance. Intuitively this makes sense; aerodynamic conductance will control how dominant balances at the land surface are communicated to the boundary layer. While $\Delta$ controlls the efficiency of latent fluxes in carying heat away from the surface, it appears this is a second order term, relative to $g_a$, for scaling $\frac{\partial \; ET}{\partial \; VPD}$.
